---
title: "Telcom Customer Churn"
author: "Nirmal Patel"
output:
  pdf_document: default
  word_document: default
  html_document: default
---


###Dataset: <https://www.ibm.com/communities/analytics/watson-analytics-blog/guide-to-sample-datasets/>
    
***
##Problems at Hand
*What is the problem I want to solve?*    
The 21st century where entertainment and communication through various devices has become a necessity, each telecom company need to know how to attract new customers as well as retain customers. In an environment where competetion is cut throat it is crucial to look at previous data and anlyze it to find the trends and predictions to see how to reduce the customer churn rate.
    
***
##Client and Purpose of Capstone
*Who is my client and why is this problem important to them? what will my client have to do or decide based on my analysis that they wouldn't have otherwise?*    
The clients range from CEO's at various telemarketing companies who want maximum business for their company to marketing firms that want to know what factors to target in order to reduce churn rates in their new campaign. The purpose of this capstone would be to take the data provided by IBM and break it down to answer questions about clientele and what variables are likely to help retain a customer.Thus generating a loyal customer base which helps revenue.
    
***    
##Data
*What data will I use for this? How will this data be acquired?*    
The data being used for this is generated by IBM for educational purposes about a telecom company which sells data plans as well as internet and television. This data is being used to evaluate which plans generate  the most sales and also how and possibly if we can minimize custommers from churning.
    
The dataset consists of 147,924 entries with 7044 rows and 21 columns. The column variables and their descriptions are:
     
|Variable       |Description    |
|-------------------------|----------------------------------------------------------------------------------|
|customerID| Customer ID
|genderCustomer| Gender (female, male)
|SeniorCitizen |Whether the customer is a senior citizen or not (1, 0)
|PartnerWhether |The customer has a partner or not (Yes, No)
|DependentsWhether |The customer has dependents or not (Yes, No)
|tenure |Number of months the customer has stayed with the company
|PhoneService|Whether the customer has a phone service or not (Yes, No)
|MultipleLines|Whether the customer has multiple lines or not (Yes, No, No phone service)
|InternetService|Customer's internet service provider (DSL, Fiber optic, No)
|OnlineSecurity|Whether the customer has online security or not (Yes, No, No internet service)
|OnlineBackup|Whether the customer has online backup or not (Yes, No, No internet service)
|DeviceProtection|Whether the customer has device protection or not (Yes, No, No internet service)
|TechSupport|Whether the customer has tech support or not (Yes, No, No internet service)
|StreamingTV|Whether the customer has streaming TV or not (Yes, No, No internet service)
|StreamingMovies|Whether the customer has streaming movies or not (Yes, No, No internet service)
|Contract|The contract term of the customer (Month-to-month, One year, Two year)
|PaperlessBilling|Whether the customer has paperless billing or not (Yes, No)
|PaymentMethod|The customer's payment method (Electronic check, Mailed check, Bank transfer (automatic), Credit card (automatic))
|MonthlyCharges|The amount charged to the customer monthlyl
|TotalCharges|The total amount charged to the customer
|ChurnWhether| The customer churned or not (Yes or No)
    
```{r telecom, echo=FALSE, include=FALSE}

library(knitr)
library(readr)
library(dplyr)


telecom <- read.csv("WA_Fn-UseC_-Telco-Customer-Churn.csv", header=T)

```
```{r telecom preview, echo=FALSE, include=TRUE}
head(telecom)
```


 
****     
##Outline
1.Find Data on Customers   
2.Clean the data using dplyr and tidyr   
3.Analyze the data using statistics   
4.Add plots and graphs to make the data visually appealing and informative    
5.Use a predictive model to find if the customer is likely to churn based on the data   
6.Organize and notate your analysis into an easily understandable report    
    
***
##Deliverables
*What are your deliverables?*     
My deliverables will consist of a report on my findings, a slide deck, and R code used to analyze the data. 

